The demographic costs of sexually antagonistic selection in partially selfing populations

### 2022-01-22 doc/SA-Hermaphrodites-wDemography-R1.tex
Update ms with Dryad info

% ALTERNATIVE TITLES:
%Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

% Population viability affects/impacts the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

 % Demographic viability of antagonistic alleles impacts conditions for the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

 %Conditions for the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites impacted by considering population viability

%\author{Colin Olito$^{1,\ast}$ \\ 
%Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

%\noindent{} 1. Department of Biology, Lund University, Lund 223 62, Sweden;

%\noindent{} 2.  Department of Evolutionary Biology and Environmental Studies, University of Zurich, Zurich CH-8057, Switzerland;

%\noindent{} $\ast$ Corresponding author; e-mail: colin.olito@gmail.com

@@

@@

\textcolor{blue}{When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other.} Classic population genetics theory has been fundamental to understanding \textcolor{blue}{how and when SA genetic variation can be maintained by balancing selection}, but these models have rarely considered the demographic consequences \textcolor{blue}{of coexisting alleles with deleterious fitness effects in each sex.} In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

%\textcolor{blue}{A sexually antagonistic polymorphism is a polymorphism of two alleles with antagonistic effects on the two sexes, that is, one of the two coexisting allele confers a fitness benefit to females and reduces male fitness, and the other allele confers a fitness benefit to males and reduces female fitness. }

%\section*{Alternative Abstract -- currently 215/200 word max.}
%%%%%%%%%%%%%%%%%%%%
%Genetic trade-offs between male and female fitness impose fascinating genetic constraints on adaptation. On one hand they can maintain genetic variation for a variety of life-history and gender-related traits, while on the other hand they cause unavoidable maladaptation as individuals are prevented from reaching their sex-specific fitness optima. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of sexually antagonistic (SA) selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained {\itshape and} the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages. Finally, we show an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

%To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1937}, \textcolor{blue}{rapid, unpredictable, or extreme changes} in environmental conditions \citep{Maynard-Smith1976, LandeShannon1996, OrrUnckless2008,Lyberger-etal-2021}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when \textcolor{blue}{different fitness components are negatively correlated} \citep{CharlesworthHughes2000, ConnallonHall2018}. 
\textcolor{blue}{A population's ability to adapt to its environment and persist in the long term depends upon the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.}

Genetic constraints on adaptation \textcolor{blue}{due to conflicting selection }have particularly interesting evolutionary consequences. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which shapes a population's capacity for future adaptation \citep{Fisher1930, Prout2000, CharlesworthHughes2000, ConnallonHall2018}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy, harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

%\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection and gene flow} have particularly interesting evolutionary consequences \citep{ConnallonDebarreLi2018}. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects on male and female sex functions can experience SA selection in hermaphrodite populations, where individual fitness is determined jointly by maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, sex-differences in directional selection are commonly detected in phenotypic selection studies \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, suggesting that many genetic variants contributing to quantitative traits have SA effects. Moreover, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in several recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA \textcolor{blue}{genetic variation} in both dioecious \textcolor{blue}{(separate-sexed)} and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.}

% Sexually antagonistic selection (abbreviated SA hereafter) \textcolor{blue}{is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{For instance, phenotypic selection studies commonly find sex-differences in directional selection \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Likewise, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

% Alleles with opposing fitness effects through male and female sex functions can also cause SA selection in hermaphrodite populations, where individuals' overall fitness is determined jointly by their maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. \textcolor{blue}{In obligately outcrossing hermaphrodite populations, SA selection operates similarly as in seperate-sexed ones (dioecious species). However, the possibility of self-fertilization has important consequences for the maintenance of SA genetic variation in hermaphrodites}. Self-fertilization reduces \textcolor{blue}{opportunities for} balanced SA polymorphisms, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014,Glemin2021}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, antagonism between viability and fecundity selection \citep{Glemin2021}, or spatial heterogeneity and complexity of the life-cycle \textcolor{blue}{can all facilitate the maintenance of} SA polymorphisms \citep{Olito-etal-2018,ConnallonSharmaOlito2019, Glemin2021}. Overall, both theoretical predictions and current empirical data suggest that there is ample scope for SA \textcolor{blue}{fitness variation} and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, DelphHerlihy2012, WangBarrett2020, RuzickaESEB2020, RuzickaConnallon2022}.

\textcolor{blue}{While significant efforts have been made to predict when SA polymorphisms can be maintained and to detect SA genetic variation from empirical data, the demographic consequences of SA selection have often been overlooked \citep{MatthewsConnallon2019}}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative trait evolution \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account to fully understand how genetic variation for fitness is shaped in natural populations \citep{metcalf2007evolutionary,coulson2006putting,rueffler2006evolutionary}.}

%a fuller understanding of the genetic basis of fitness variation, and the processes maintaining it, requires an explicit treatment of the demographic processes involved in evolutionary change.

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with SA selection and population dynamics, but did not analyze how demography influenced SA polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to \textcolor{blue}{include both dioecious and hermaphroditic populations and jointly analyze the genetic (fixation/polymorphism) and demographic (persistence/extinction) outcomes of SA selection}. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

% \bigskip

% \begin{enumerate}
% 	\item Introduce Evolutionary Demography, pivot to focus on the consequences of taking demography into account explicitly for the maintenance of SA polymorphism. [Lotte]
% 	\item Introduce key difference between existing population genetic models of SA polymorphism and our models: possibility of diverse population and evolutionary dynamics. [Lotte then Colin]
% 	\item Explain and highlight "demographically viable" parameter space, indicate that we are interested in whether parameter space predicted to be polymorphic by Pop.~gen.~models is, indeed, "demographically viable" [Lotte then Colin]
% 		\begin{enumerate}
% 			\item Pop. Gen. models tacitly assume large and/or stable population size, parameterize genotypic fitnesses relative to the 'best performing' genotype. Non-overlapping generations. Casually define fitness as lifetime reproductive fitness of a given genotype.
% 			\item "In practice, these genotypic fitnesses emerge from myriad process acting throughout the life-history of individuals"
% 			\item these processes contribute to key demographic rates such as the intrinsic population growth rate, 
% 		\end{enumerate}
% 		\item A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations.
% \end{enumerate}
% \bigskip

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. \textcolor{blue}{We focus our presentation on a two-stage example of the model and discuss its' underlying biological assumptions. The derivation of the general model for any number of age or stage classes is provided in the Online Supplementary Material.} The derivation follows closely the logic presented in \citet{deVriesCaswell2019a, deVriesCaswell2019b}, and in fact, it reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing \textcolor{blue}{(when selfing is not allowed)}. All computer code necessary to reproduce the results are available at \url{https://anonymous.4open.science/r/SA-Hermaphrodites-wDemography-4FFF}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and can potentially reproduce by a combination of self- and outcross-fertilization. Maternal outcrossing involves receiving male gametes from other individuals in the population, while paternal outcrossing involves exporting male gametes to fertilize other individuals' ovules. Self-fertilization occurs when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female reproductive functions, we denote matrices or vectors relating to the male sex function with a prime. \textcolor{blue}{Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen. We present a more general model of genotype-specific self-fertilization rates in the Online Supplement, and briefly discuss implications of this approach in the Discussion.}

Individuals are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively; see Table \ref{tab:Terms} for a full description of terms included in the model). We include how individuals were produced in the state description because individuals produced by selfing might experience reduced survival, growth, or maturation rates due to inbreeding depression. \textcolor{blue}{This approach makes the common simplifying assumption that all individuals in the population experience the same severity of inbreeding depression (e.g., \citealt{Charlesworth1987, Charlesworth2009, Charlesworth2010, JordanConnallon2014}). We address potential consequences of relaxing this assumption, and how our model can be extended to do so, in the discussion. }

@@

Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ g\\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$g \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times 2\omega g$ \\
\hline
\multicolumn{3}{l}{Parameters used in the two stage example} \\
Symbol & Definition & Value/Range\\
\hline
$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & 0-0.15\\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & $\frac{1}{2}$ or $\frac{1}{4}$\\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & 0.6\\
$\gamma$ & Transition rate from juvenile to adult stages & 0.05\\
$C$ & The population rate of self-fertilization &0-1 \\
$\delta$ &\textcolor{blue}{ Inbreeding depression effect on zygote viability}&0-0.8\\
$ \delta_j$ & \textcolor{blue}{ Inbreeding depression effect on juvenile survival}& 0-0.8\\
$\delta_a$ & \textcolor{blue}{ Inbreeding depression effect on adult survival}& 0-0.8\\
$\delta_{\gamma}$ &\textcolor{blue}{Inbreeding depression effect on maturation rate} & 0-0.8\\
$f_i,\,f^{\prime}_i$ & Adult \textcolor{blue}{maximum} fertilities through female and male sex functionsxi & 5.8-8.5 \\

@@

\textcolor{blue}{If the survival or fertility rates are a function of the population density, $\tilde{\mbf{n}}(t)$, then the model will be density-dependent and the population projection matrix becomes a function of the full population vector, $\tilde{\mbf{A}}[\tilde{\mbf{n}}(t)]$. Below, we limit our analyses to consider density-independent survival and fertility rates. The population projection matrix is then only a function of the proportional population vector, $\tilde{\mbf{A}}[\tilde{\mbf{p}}(t)]$, and the model is referred to as frequency-dependent. In \ref{App:DensDep} we construct and analyze a simple density-dependent version of the model and show that with density-dependent survival affecting all stages equally, our main conclusions remain intact.}

@@

\textcolor{blue}{We now construct and analyze an example genotype $\times$ stage-classified model for a hypothetical hermaphrodite species with intralocus sexual conflict via the two sex functions.} For the sake of simplicity, we assume our hypothetical species has a life cycle with only two stages: juveniles and adults (i.e., $\omega = 2$), and that only adults can reproduce. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

@@

\noindent where $f$ and $f^{\prime}$ represent \textcolor{blue}{maximum adult fertilities}, and $w_{i}$ and $w^{\prime}_{i}$ the genotypic relative scaling factors for female and male sex-functions (see Table \ref{tab:Fitness}).

>>>>

\author{Colin Olito$^{1,\ast}$ \\ 
Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

\noindent{} 1. Department of Biology, Lund University, Lund 223 62, Sweden;

\noindent{} 2.  Department of Evolutionary Biology and Environmental Studies, University of Zurich, Zurich CH-8057, Switzerland;

\noindent{} $\ast$ Corresponding author; e-mail: \url{colin.olito@gmail.com}

@@

\textit{Online enhancements}: Dryad digital repository ({\itshape https://doi.org/10.5061/dryad.c2fqz619t}).

@@

When selection differs between the sexes, genes expressed by both males and females can experience sexually antagonistic (SA) selection, where beneficial alleles for one sex are deleterious for the other. Classic population genetics theory has been fundamental to understanding how and when SA genetic variation can be maintained by balancing selection, but these models have rarely considered the demographic consequences of coexisting alleles with deleterious fitness effects in each sex. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (i.e., dioecious/gonochorous) and partially selfing hermaphrodite populations. We focus on identifying when SA polymorphisms are maintained by balancing selection {\itshape and} the population growth rate remains positive. Additionally, we analyze the effects of inbreeding depression manifesting at different life-history stages and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results show that when population intrinsic growth rates approach one, extinction occurs across large swathes of parameter space favoring SA polymorphism or the fixation of male-beneficial alleles, and that inbreeding depression is a significant problem for maintaining SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations can sustain large regions of viable SA polymorphic space.

@@

A population's ability to adapt to its environment and persist in the long term depends upon the genetic variation that it carries \citep{Fisher1930, Lewontin1975}. Yet, populations are heterogenous collections of individuals existing in a complex world where adaptation can be impeded by conflicting selection and gene-flow between different classes of individuals \citep{Prout2000, CharlesworthHughes2000, ConnallonHall2018}.

Genetic constraints on adaptation due to conflicting selection have particularly interesting evolutionary consequences. On one hand, they provide an effective mechanism for the maintenance of genetic variation, which shapes a population's capacity for future adaptation \citep{Fisher1930, Prout2000, CharlesworthHughes2000, ConnallonHall2018}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy, harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) is an important form of conflicting selection that occurs when beneficial alleles for one sex are deleterious when expressed in the other. Likewise, alleles with opposing fitness effects on male and female sex functions can experience SA selection in hermaphrodite populations, where individual fitness is determined jointly by maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. SA selection appears to be a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. For instance, sex-differences in directional selection are commonly detected in phenotypic selection studies \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, suggesting that many genetic variants contributing to quantitative traits have SA effects. Moreover, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}), and polygenic signals of SA selection are evident in several recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}. Overall, both theory and current empirical data suggest that there is ample scope for the maintenance of SA genetic variation in both dioecious (separate-sexed) and hermaphrodite populations \citep{BondurianskyChenoweth2009, Abbott2011, RuzickaESEB2020, WangBarrett2020}.

While significant efforts have been made to predict when SA polymorphisms can be maintained and to detect SA genetic variation from empirical data, the demographic consequences of SA selection have often been overlooked \citep{MatthewsConnallon2019}. Most population genetic theory, including SA selection theory, models the relative fitness of genotypes in populations of constant (and often infinite) size. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals that link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). For over 50 years, the field of eco-evolutionary dynamics has sought to jointly model the genetic and demographic processes involved in evolutionary change. Theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic consequences of allele frequency changes due to selection. Various extensions of these theories have been made to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative trait evolution \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects}. This growing body of eco-evolutionary theory makes plain that demographic processes must be taken into account to fully understand how genetic variation for fitness is shaped in natural populations \citep{metcalf2007evolutionary,coulson2006putting,rueffler2006evolutionary}.

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with SA selection and population dynamics, but did not analyze how demography influenced SA polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic populations and jointly analyze the genetic (fixation/polymorphism) and demographic (persistence/extinction) outcomes of SA selection. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. We focus our presentation on a two-stage example of the model and discuss its' underlying biological assumptions. The derivation of the general model for any number of age or stage classes is provided in the Online Supplementary Material. The derivation follows closely the logic presented in \citet{deVriesCaswell2019a, deVriesCaswell2019b}, and in fact, it reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing (when selfing is not allowed). All computer code and data necessary to reproduce the results are available on Dryad (\url{https://doi.org/10.5061/dryad.c2fqz619t}) and on GitHub (\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}).

Simultaneous hermaphrodites can transmit genes to the next generation via both sperm/pollen and eggs/ovules, and can potentially reproduce by a combination of self- and outcross-fertilization. Maternal outcrossing involves receiving male gametes from other individuals in the population, while paternal outcrossing involves exporting male gametes to fertilize other individuals' ovules. Self-fertilization occurs when an individual's male gametes fertilize their own ovules. To distinguish between parameters relating to male and female reproductive functions, we denote matrices or vectors relating to the male sex function with a prime. Following previous theory for partially selfing populations \cite[e.g.,][]{Charlesworth2010,JordanConnallon2014,Glemin2021}, we use a 'fixed prior selfing' model where all individuals are assumed to self-fertilize at a constant rate, $C$, prior to receiving outcross pollen. We present a more general model of genotype-specific self-fertilization rates in the Online Supplement, and briefly discuss implications of this approach in the Discussion.

Individuals are jointly classified by life-cycle stage ($1, \ldots, \omega$), genotype ($1, \ldots, g$), and whether they were produced by self- or outcross fertilization (denoted by $S$ and $X$ superscripts, respectively; see Table \ref{tab:Terms} for a full description of terms included in the model). We include how individuals were produced in the state description because individuals produced by selfing might experience reduced survival, growth, or maturation rates due to inbreeding depression. This approach makes the common simplifying assumption that all individuals in the population experience the same severity of inbreeding depression (e.g., \citealt{Charlesworth1987, Charlesworth2009, Charlesworth2010, JordanConnallon2014}). We address potential consequences of relaxing this assumption, and how our model can be extended to do so, in the discussion.

@@

Symbol & Definition & Dimension \\
\hline
$g$      & Number of genotypes ($3$; $AA$, $Aa$, and $aa$) & \\
$\omega$ & Number of stages ($2$; juvenile and adult) & \\
$\tilde{\mbf{n}}$ & Joint stage $\times$ genotype vector & $2 \omega g \times 1$ \\
$\tilde{\mbf{p}}$ & Joint stage $\times$ genotype frequency vector & $2 \omega g \times 1$ \\
$\mbf{U}^{S}_{i}$,\, $\mbf{U}^{X}_{i}$ & Genotype-specific transition and survival matrices & $\omega \times \omega$ \\
$\mcal{U}^{S},\, \mcal{U}^{X}$ & Block diagonal selfed/outcrossed survival matrices & $\omega g \times \omega$ g\\
$\mbf{F}_{i}$,\, $\mbf{F}^{\prime}_{i}$ & Genotype-specific fertility matrices & $\omega \times \omega$ \\
$\mcal{F}^{S},\, \mcal{F}^{X}$ & Block diagonal selfed/outcrossed fertility matrix & $\omega g \times \omega$g \\
% $\mbf{K}$   & Vec-permutation matrix & $g \times g$ \\
% $\mbf{I}$   & Identity matrix & given in subscript \\
$q^{\prime}_{A}$,\, $q^{\prime}_{a}$ & Allele frequencies in male gamete pool & $1$ \\
$\tilde{\mbf{A}}(\tilde{\mbf{p}})$ & Population projection matrix & $2 \omega g \times 2\omega g$ \\
\hline
\multicolumn{3}{l}{Parameters used in the two stage example} \\
Symbol & Definition & Value/Range\\
\hline
$s_f,\,s_m$ & Selection coefficients through female and male reproductive function & 0-0.15\\
$h_f,\,h_m$ & Dominance coefficients through female and male reproductive function & $\frac{1}{2}$ or $\frac{1}{4}$\\
$\sigma_j,\, \sigma_a$ & Survival rates for juvenile and adult stages & 0.6\\
$\gamma$ & Transition rate from juvenile to adult stages & 0.05\\
$C$ & The population rate of self-fertilization &0-1 \\
$\delta$ & Inbreeding depression effect on zygote viability&0-0.8\\
$ \delta_j$ &  Inbreeding depression effect on juvenile survival & 0-0.8\\
$\delta_a$ &  Inbreeding depression effect on adult survival & 0-0.8\\
$\delta_{\gamma}$ &Inbreeding depression effect on maturation rate & 0-0.8\\
$f_i,\,f^{\prime}_i$ & Adult maximum fertilities through female and male sex functions & 5.8-8.5 \\

@@

If the survival or fertility rates are a function of the population density, $\tilde{\mbf{n}}(t)$, then the model will be density-dependent and the population projection matrix becomes a function of the full population vector, $\tilde{\mbf{A}}[\tilde{\mbf{n}}(t)]$. Below, we limit our analyses to consider density-independent survival and fertility rates. The population projection matrix is then only a function of the proportional population vector, $\tilde{\mbf{A}}[\tilde{\mbf{p}}(t)]$, and the model is referred to as frequency-dependent. In \ref{App:DensDep} we construct and analyze a simple density-dependent version of the model and show that with density-dependent survival affecting all stages equally, our main conclusions remain intact.

@@

We now construct and analyze an example genotype $\times$ stage-classified model for a hypothetical hermaphrodite species with intralocus sexual conflict via the two sex functions. For the sake of simplicity, we assume our hypothetical species has a life cycle with only two stages: juveniles and adults (i.e., $\omega = 2$), and that only adults can reproduce. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect. Following convention, we parameterize the fertility component of fitness for each genotype through each sex function, $w_{i}$ and $w^{\prime}_{i}$, to be bounded by $[0,1]$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}).

@@

\noindent where $f$ and $f^{\prime}$ represent maximum adult fertilities, and $w_{i}$ and $w^{\prime}_{i}$ the genotypic relative scaling factors for female and male sex-functions (see Table \ref{tab:Fitness}).

Unless stated otherwise, we use the following parameter values for the demographic rates in the model: $\sigma_j = \sigma_a = 0.6$ and $\gamma = 0.05$. These chosen values are similar to those used in \citet{deVriesCaswell2019b}, facilitating comparison between models, and correspond to a life-history in which individuals spend multiple timesteps in the juvenile phase prior to maturing into reproductively active adults, but are otherwise arbitrary. Our parameters of interest include maximum female fertility, $f$, the inbreeding depression parameters, $\delta$, $\delta_j$, $\delta_a$, $\delta_{\gamma}$, and the selection parameters $h_f$, $s_f$, $h_m$, and $s_m$, which are given different values for each analyses as described in the figure captions.
